---
title: "HW1"
author: "chandl71@purdue.edu"
date: "2024-06-16"
output: html_document
---

```{r setup}
le <- read.csv("./life_expectancy.csv")
X <- le$Medical.doctors
Y <- le$X2015Life.expectancy
```

# Problem 1

Estimate the parameters ($\beta_0$, and $\beta_1$) for a linear regression to predict $Y$ based on $X$. Complete the following with details.

## 1.a

```{r}
Xbar <- mean(X)
```

$\bar{X} = `r Xbar`$

## 1.b

```{r}
Ybar <- mean(Y)
```

$\bar{Y} = `r Ybar`$

## 1.c

```{r}
sum_mean_diff_xy = sum((X - Xbar) * (Y - Ybar))
```

$\sum(X_i - \bar{X})(Y_i - \bar{Y}) = `r sum_mean_diff_xy`$

## 1.d

```{r}
sum_mean_sq_diff_x = sum((X - Xbar)^2)
```

$\sum(X_i - \bar{X})^2 = `r sum_mean_sq_diff_x`$

## 1.e

Start with $b_1$ calculation since $b_0$ can be easily calculated from there:

$$
b_1 = \frac{\sum(X_{i} - \bar{X})(Y_{i} - \bar{Y})}{\sum(X_{i}-\bar{X})^{2}}
$$

```{r}
b1 = sum_mean_diff_xy / sum_mean_sq_diff_x
```

Now calculate $b_0$ using $b_1$:

$$
b_{0} = \bar{Y} - b_{1} \bar{X}
$$

```{r}
b0 = Ybar - b1 * Xbar
```

$b_0 = `r b0`$

## 1.f

Using previously calculated value for $b_1$:

$b_1 = `r b1`$

## 1.g

First, calculate $\hat{Y}$:

$$
\mathbf{\hat{Y}} = b_0 + b_1 * \mathbf{X}
$$

``` {r}
Yhat <- b0 + b1 * X
```

then proceed with $SSE$:

```{r}
sse = sum((Y - Yhat)^2)
```

$SSE = \sum(Y_i - \hat{Y})^2 = `r sse`$

## 1.h

```{r}
n = length(X)
dfe = n - 2
```

We have `r dfe` degrees of freedom of error, and we already know $SSE$, so

```{r}
mse = sse / dfe
```

$MSE = \frac{SSE}{DFE} = `r mse`$

## 1.i
```{r}
sst = sum((Y - Ybar)^2)
```

$SSE = \sum(Y_i - \bar{Y})^2 = `r sst`$

## 1.j
```{r}
ssr = sum((Ybar - Yhat)^2)
```

$SSR = \sum(\bar{Y} - \hat{Y}_i)^2 = `r ssr`$

To determine whether the two sides of the equation $SST = SSR + SSE$ are equal, we will compare their absolute difference against a very small threshold:

```{r}
thresh = 1e-10
abs(sst - (ssr + sse)) < thresh
```

Since the above is true, the values are close enough for us to consider them equal.

# Problem 2

Obtain t-values at $\alpha = 0.1, 0.05$ using the $n - 2 = `r dfe`$ degrees of freedom for this problem:

``` {r}
t_0.1 = qt(1 - 0.1 / 2, df = dfe)
t_0.05 = qt(1 - 0.05 / 2, df = dfe)
```

Next, obtain the standard error of the estimation $s{b_1}$ and use it to obtain the margin of error for each $t$ value:

``` {r}
stderr_b1 = sqrt(mse / sum_mean_sq_diff_x)
me_0.1 = t_0.1 * stderr_b1
me_0.05 = t_0.05 * stderr_b1
```

In order to estimate the linear impact of $X$ on $Y$, at a confidence of $(1 − \alpha)%$, you should use the critical
value, or the $t$ value denoted as $t(1 - \alpha / 2, n - 2)$, which has a value of `r t_0.1` at $\alpha = 0.1$, and `r t_0.05` at $\alpha = 0.05$. The standard error of the estimation is:
$$
s{b_1} = \sqrt{\frac{MSE}{\sum(X_i - \bar{X})^2}} = `r stderr_b1`
$$
The margin error, or $t \cdot SE$, of the confidence interval is `r me_0.1` at $\alpha = 0.1$, and `r me_0.05` at $\alpha = 0.05$.

# Problem 3
Perform a hypothesis test on the linear impact of $X$ on $Y$, with a $t$ test with a significant value of $0.1$.

TODO DEFINE H0 Ha

We assume the use of a two-sided test against 0, and we define our test statistic as:

$$
t^\star = \frac{(b_1 - 0)}{s{b_1}} 
$$

```{r}
t_star = b1 / stderr_b1
```

$t^\star = `r t_star`$, so we need to compare this to the $t$ value for $\alpha = 0.1$, which was calculated in the previous problem as $t(1 - \frac{0.1}{2}, n - 2) = `r t_0.1`$.

Since $t^\star = `r t_star` > `r t_0.1`$