---
title: "HW7"
author: "Robert Chandler"
email: "chandl71@purdue.edu"
date: "2024-07-28"
output:
  pdf_document: default
---

# Problem 1

Consider a model to predict the average length of a physician’s stay based on
the four regions. The one-way ANOVA model $\log_10Y \sim \text{Region}$ can be
represented with different factor effects model, depending on the baseline.
Hence, the mean responses, $\mu_1$, $\mu_2$, $\mu_3$, and $\mu_4$ are the means
computed on $\log_10Y$.

Build a model summary using the first region (1) as the baseline, the R output
shows table as the following. The values in the corresponding cells are denoted
by $A_i, \quad i = \{1, 2, ..., 16\}$.

## 1.a

Based on Table 1, The means, $\mu_1$, $\mu_2$, $\mu_3$, and $\mu_4$ can be
represented as:

$$
\begin{aligned}
\mu_1&: A_1 + 0 = A_1 \\
\mu_2&: A_1 + A_2 \\
\mu_3&: A_1 + A_3 \\
\mu_4&: A_1 + A_4 \\
\end{aligned}
$$

## 1.b

Based on Table 2, The means, $\mu_1$, $\mu_2$, $\mu_3$, and $\mu_4$ can be
represented as:

$$
\begin{aligned}
\mu_1&: B_1 + B_2 \\
\mu_2&: B_1 + 0 = B_1 \\
\mu_3&: B_1 + B_3 \\
\mu_4&: B_1 + B_4 \\
\end{aligned}
$$

## 1.c 

To perform the hypothesis $H_0: \mu_2 = \mu_1$, the point estimate $\hat{Y}_2 -
\hat{Y}_1 = \mu_2 - \mu_1 = A_1 + A_2 - A_1 = A_2$ and the p-value is $A_{14} =
B_{14}$.

## 1.d 

To perform the hypothesis $H_0: \mu_3 = \mu_1$, the point estimate $\hat{Y}_3 -
\hat{Y}_1 = \mu_3 - \mu_1 = A_1 + A_3 - A_1 = A_3$ and the p-value is $A_{15}$.

# Problem 2

According to the following data summary, compute the following confidence
intervals:

```{r}
df <- data.frame(
  group = c(1, 2, 3, 4), mean = c(25, 46, 18, 62), n = c(108, 103, 152, 77)
)
```

## 2.a

The difference between the average $Y$ in Groups 1 and 2 and in Groups 3 and 4, i.e.

$$
\frac{\mu_1 + \mu_2}{2} - \frac{\mu_3 + \mu_4}{2}
$$

```{r}
alpha <- 0.05
n <- sum(df$n)
r <- length(unique(df$group))
mse <- 100
t_crit <- qt(1 - alpha / 2, n - r)
l_hat <- 0.5 * (sum(df$mean[1:2]) - sum(df$mean[3:4]))
stderr <- sqrt(mse * 0.5^2 * sum(1 / df$n))
interval_lower <- l_hat - stderr * t_crit
interval_upper <- l_hat + stderr * t_crit
```

$$
L = \frac{\mu_1 + \mu_2}{2} - \frac{\mu_3 + \mu_4}{2}
$$

so $c_1 = \frac{1}{2}, c_2 = \frac{1}{2}, c_3 = -\frac{1}{2}, c_4
= - \frac{1}{2}$.

$$
\hat{L} = \frac{\bar{Y}_1 + \bar{Y}_2}{2} - \frac{\bar{Y}_3 + \bar{Y}_4}{2}
= \frac{`r df$mean[1]` + `r df$mean[2]`}{2} - \frac{`r df$mean[3]` + `r df$mean[4]`}{2} 
= `r l_hat`
$$

Using the $MSE$ and $n_i$ from the table:

$$
s^2[L] = MSE \sum_{i=1}^r \frac{c_i^2}{n_i} = 100\left(\frac{(1/2)^2}{108} +
\frac{(1/2)^2}{103} + \frac{(-1/2)^2}{152} + \frac{(-1/2)^2}{77}\right)
= `r stderr^2`
$$

so

$$
s[\hat{L}] = `r stderr`
$$

and with $df_E = n_T - r = 108 + 103 + 152 + 77 - 4 = `r n - r`$, the critical
t-value is:

$$
t(1 - \alpha / 2, n - r) = t(`r 1 - alpha / 2`, `r n - r`) = `r t_crit`
$$

Therefore, the 95% confidence interval on the difference is:

$$
CI = \hat{L} \pm t(1 - \alpha / 2, n - r) s[\hat{L}] = (`r c(interval_lower, interval_upper)`)
$$

## 2.b

The simultaneous confidence intervals for the differences of the following:

$$
\mu_1 - \mu_2, \quad \mu_1 - \mu_3, \quad \mu_2 - \mu_3
$$

```{r}
group_names <- c("1-2", "1-3", "2-3")
g <- length(group_names)
bonf <- qt(1 - alpha / (2 * g), n - r)
df_bonf <- data.frame()
coeffs <- list(
  c(1, -1, 0, 0),
  c(1, 0, -1, 0),
  c(0, 1, -1, 0)
)

df_bonf[group_names, "l_hat"] <- sapply(
  coeffs,
  function(coef) sum(df$mean * coef)
)

df_bonf[group_names, "stderr"] <- sapply(
  coeffs,
  function(coef) sqrt(mse * sum(coef^2 / df$n))
)

intervals <- t(rbind(
  df_bonf$l_hat - bonf * df_bonf$stderr,
  df_bonf$l_hat + bonf * df_bonf$stderr
))
rownames(intervals) <- group_names
colnames(intervals) <- c(
  paste(100 * alpha / 2, "%"), paste(100 * (1 - alpha / 2), "%")
)
```

Calculating the critical Bonferroni value:

$$
B = t(1 - \alpha / (2g), n_T - r)
$$

The $\hat{L}$ for each case are:

$$
\begin{aligned}
\bar{Y}_1 - \bar{Y}_2 &= `r df$mean[1]` - `r df$mean[2]` = `r df_bonf["1-2", "l_hat"]` \\
\bar{Y}_1 - \bar{Y}_3 &= `r df$mean[1]` - `r df$mean[3]` = `r df_bonf["1-3", "l_hat"]` \\
\bar{Y}_2 - \bar{Y}_3 &= `r df$mean[2]` - `r df$mean[3]` = `r df_bonf["2-3", "l_hat"]` \\
\end{aligned}
$$

The $s[\hat{L}]$ for each case are:

$$
\begin{aligned}
\sqrt{MSE \sum_{i = 1}^r c_i^2 / n_i} &= \sqrt{`r mse` \left(\frac{1}{`r df$n[1]`} + \frac{1}{`r df$n[2]`}\right)} = `r df_bonf["1-2", "stderr"]` \\
\sqrt{MSE \sum_{i = 1}^r c_i^2 / n_i} &= \sqrt{`r mse` \left(\frac{1}{`r df$n[1]`} + \frac{1}{`r df$n[3]`}\right)} = `r df_bonf["1-3", "stderr"]` \\
\sqrt{MSE \sum_{i = 1}^r c_i^2 / n_i} &= \sqrt{`r mse` \left(\frac{1}{`r df$n[2]`} + \frac{1}{`r df$n[3]`}\right)} = `r df_bonf["2-3", "stderr"]` \\
\end{aligned}
$$

And using the formula

$$
\hat{L} \pm Bs[\hat{L}]
$$

for each group

$$
\begin{aligned}
`r df_bonf["1-2", "l_hat"]` &\pm `r bonf` \cdot `r df_bonf["1-2", "stderr"]` \\
`r df_bonf["1-3", "l_hat"]` &\pm `r bonf` \cdot `r df_bonf["1-3", "stderr"]` \\
`r df_bonf["2-3", "l_hat"]` &\pm `r bonf` \cdot `r df_bonf["2-3", "stderr"]` \\
\end{aligned}
$$

the intervals are

$$
\begin{aligned}
\mu_1 - \mu_2&: (`r intervals["1-2", ]`) \\
\mu_1 - \mu_3&: (`r intervals["1-3", ]`) \\
\mu_2 - \mu_3&: (`r intervals["2-3", ]`) \\
\end{aligned}
$$
