---
title: "HW3"
author: "Robert Chandler"
email: "chandl71@purdue.edu"
date: "2024-06-30"
output:
  pdf_document: default
---

# Setup
Begin by reading the dataset for this set of problems:

```{r setup}
le <- read.csv("../../datasets/life_expectancy.csv")
```

```{r}
y <- le$X2015Life.expectancy
x1 <- le$Medical.doctors
x2 <- le$Nurses
x3 <- le$Pharmacists
```

Utilize the life expectancy data and examine a simple linear regression $Y \sim X$,
where $X = X_3$, and $Y$ represents life expectancy.

Set $X$ accordingly:

```{r}
x <- x3
```

# Problem 1

At a confidence level of 95%...

# 1.a

```{r}
alpha <- 0.05
n <- length(y)
t_crit <- qt(1 - alpha / 2, n - 2)
b <- qt(1 - alpha / 4, n - 2)
```

To find the individual confidence intervals for $\beta_0$, the critical value is denoted by
$t(1 - \alpha / 2, n-2)$ and has a value of $t(1 - `r alpha` / 2, `r n` - 2) =
`r t_crit`$.

# 1.b

To find the individual confidence intervals for $\beta_1$, the critical value is denoted by
$t(1 - \alpha / 2, n-2)$ and has a value of $t(1 - `r alpha` / 2, `r n` - 2) =
`r t_crit`$.

# 1.c

To calculate the Bonferroni joint confidence intervals for all parameters using
an $\alpha = 0.05$, the critical value is denoted by $t(1- \frac{\alpha}{2p},
DFE ) = t(1- \frac{\alpha}{4}, n - 2)$ and has a value of $t(1 - \frac{`r
alpha`}{4}, `r n` - 2) = `t_crit_bonf`$.

# 1.d

Calculate the Working-Hotelling joint confidence intervals for all parameters
using an $\alpha = 0.05$. 

```{r}
w <- sqrt(2 * qf(1 - alpha, 2, n - 2))
model <- lm(y ~ x)
model_summary <- summary(model)
model_summary
b0 <- model$coefficients[["(Intercept)"]]
b1 <- model$coefficients[["x"]]
t0 <- model_summary$coefficients["(Intercept)", "t value"]
t1 <- model_summary$coefficients["x", "t value"]

b0_interval <- c(b0 - w * t0, b0 + w * t0)
b1_interval <- c(b1 - w * t1, b1 + w * t1)
```

The interval for $b_0$ is $(`r b0_interval`)$, and the interval for $b1$ is $(`r 
b1_interval`)$

The critical value is denoted by $W = \sqrt{2 F(1 - \alpha; 2, n - 2)}$ and has
a value of $W = \sqrt{2 F(1 - \alpha; 2, n - 2)}$.

# 1.e

Which joint confidence interval is better? Briefly explain

For joint estimation of the *parameters*, the Bonferroni interval is better
because it yields a smaller critical value: $W = `r w` > `r b` = B$. This is
consistent across all models when we are estimating the
parameters/weights/coefficients simultaneously for a linear model. That is, $W/B
> 1$.

# Problem 2

At the confidence level of 95%...

## 2.a

To find the individual confidence interval for the mean response ğ‘Œà¯›
à·¢, where ğ‘‹à¯› =
the mean of the sample, the critical value is denoted by ______and has a value of ______, the standard
error can be computed with the formula ___________and has a value of ________.

## 2.b

To find the individual confidence interval for the mean response ğ‘Œà¯›
à·¢, where ğ‘‹à¯› =
the median of the sample, the critical value is denoted by ______and has a value of ______, the
standard error can be computed with the formula ___________and has a value of ________.

## 2.c

To compute the joint confidence interval for the mean responses for 5 different ğ‘‹à¯› levels
(data points), the Bonferroni critical value is denoted by ______and has a value of ______; the WH
critical value is denoted by ______and has a value of _____.
Which method is better here and why?
Which method should you suggest if you are not sure what ğ‘‹ levels are to be predicted?

## 2.d

To calculate the Bonferroni joint confidence interval for the single response prediction at 3 ğ‘‹à¯›
levels, the critical value is denoted by ______and has a value of ______,

## 2.e

To calculate the Scheffe joint confidence interval for the single response predictions at 3 ğ‘‹à¯›
levels, the critical value is denoted by ______and has a value of ______,
