

## 4

Multicollinearity because the second row is much smaller in both cases

## 5 \*

$$
R^2_{Y 3 | 1,2} = \frac{SSR(X_3 | X_1, X_2)}{SSE(X_1, X_2)}
$$


```{r}
sse_x123 <- 600
ssr_x3_x12 <- 250
r_sq_3_12 <- ssr_x3_x12 / (ssr_x3_x12 + sse_x123)
rnd(r_sq_3_12)
```

## 6 \*

Too easy? CHECK IF TABLES ARE ACTUALLY TYPE I

## 7

Townhome: X2 = 1, X3 = 0

$$
\beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + 0
$$

## 8 * Semi-confident



F = MSR / MSE
= ((SSE_R - SSE_F) / (df_ER - df_EF))
/ (SSE_F / df_EF)

```{r}
dfer <- 57
dfef <- 55
sser <- 4350
ssef <- 4000
(sser - ssef) / (dfer - dfef) / (ssef / dfef)
```

## 9

Check picture

## 10

Very tough. Either all linear or X3 could be nonlinear.

## 11

```{r}
n <- 250
p <- 3
dffits <- 2 * sqrt(p / n)
cooks <- 0.50
dfbetas <- 2 / sqrt(n)
dffits
cooks
dfbetas
```

## 12

Last is only one that uses indices right

## 13

Weird because the other does interact too but we're going with the one where
they move in opposite directions because it is so clear (see pg 827 for a
discussion in the book on this being an interaction of factors)

## 14

```{r}
msr <- 250 / 4
mse <- 500 / (99 - 4)
msr / mse
```

## 15

```{r}
l_hat <- (30 + 35 + 15 + 45) / 4 - 20
cs <- c(-1, 0.25, 0.25, 0.25, 0.25)
n <- c(25, 35, 25, 40, 35)
r <- 5
stderr <- sqrt(25 * sum(cs^2 / n))
t_crit <- qt(1 - alpha / 2, sum(n) - r)
l_hat - stderr * t_crit
l_hat + stderr * t_crit
```


